Reinforcement-Learning Windturbine Controller

### 2019-12-07 paper/paper.tex
Reworked death conditions part

Machine learning has developped immensly over the last years. Benefitting from Moore's law, huge amounts of computational power can be used to train more and more complex models. Machine learning is generally split into three big branches. Supervised learning, unsupervised learning and reinforcement learning. In supervised learning, some form of fitting is done on a dataset.

@@

\section{Gym Experiments}

At the start of our experimentation, we took off with a well-known problem and a proven implementation. We used the OpenAI-Pendulum task, in which a pendulum in gravity needs to be held upright by applying a force to it. OpenAI Gyms generally serve as benchmarks for reinforcement learning algorithms and most foundational papers and articles evaluate their solutions based on how well they perform in OpenAI Gyms. We managed to find a solution with our DDPG implementation. We hoped that the implementation might yield results on our windturbine environment as well and exchanged the OpenAI gym for qblade.

@@

@@

We normalized the data so states, actions and reward usually stayed between 1 and -1. This is a normal technique in machine learning, and we are irritated by the fact that the reinforcement learning community largely made no mention of normalization. At times, rewards were clipped \cite{mnihPlayingAtariDeep2013} and observations were sometimes, especially when working on pixel inputs, cropped to be of square dimensions. Only in 2016, \cite{vanhasseltLearningValuesMany2016} brought up the topic. They propose an adaptive normalization mechanism which integrates into a generic \ac{RL} algorithm and which can normalize all inputs adaptively. We decided to implement a more straight-forward variant. At first, we used observed states and rewards from the random exploration phase to calculate the 5\%, 95\% quartile values. We calculated normalization constants from it to linearly normalize those quartiles to $[-1, 1]$. Plain min/max normalization worked less well. As sometimes during random exploration, no extreme values were observed and later extreme values lay far outside of $[-1, 1]$, we instead normalized on replay data of a previous run. We could also imagine manually setting normalization bounds on low-dimensional problems, as human knowledge about the possible state space is usually present. In case when human knowledge about input magnitudes is not present, we recommend the mechanism from \cite{vanhasseltLearningValuesMany2016}, or if you are willing to accept the extra computational effort of running once without normalization, our version.

@@

\subsection{Regarding death conditions}

In \ac{DQN}, death conditions are accounted for by changing the Q target in the terminal step to be $y = r$ instead of $y = r + \gamma \max_a' Q(s', a')$. This way, behavior after a death will not be accounted for by the Q function. As we are seeing many death conditions, we thought it could be useful to include this into our DDPG algorithm. So now, instead of $y = r + \gamma Q'(s', \mu'(s'))$ we change it to be $y = r + \gamma Q'(s', \mu'(s')) * (1-d)$ where $d \in {0, 1}$ symbolize the death conditions.

@@

\For{episode = 1, M}{
  Initialize a random process $\mathcal{N}$ for action exploration\\
  Receive initial observation state $s_1$\\
  \For{t = 1, T}{
    Select action $a_t = \mathrm{clip}(\mu(s_t;\theta^{\mu}) + \mathcal{N}_t)$ according to the current policy and exploration noise\\
    Execute action $a_t$ in $\varepsilon$ and observe $r_t$, $d_t$ and $s_{t+1}$ and normalize these\\
    Store transition with highest priority $(s_t, a_t, r_t, s_{t+1}, d_t, \max_i(p_i))$ in $\mathcal{R}$\\
    Calculate sampling probabilities $P(i) = \frac{p_i}{\sum_j p_j}$\\
    Sample a random minibatch of $N$ transitions $(s_i, a_i, r_i, s_{i+1}, d_i)$ from $\mathcal{R}$ according to $P$\\
    Calculate importance sampling weights $w_i = (N \times P(i))^{-\beta}$ and normalize $ w_i = \frac{w_i}{\max_j w_j}$\\
    Set $y_i = r_i + \gamma Q'(s_{i+1}, \mu'(s_{i+1};\theta^{\mu'});\theta^{Q'}) (1 - d_i)$\\
    Calculate TD-Error $\delta_i = y_i - Q(s_i, a_i;\theta^Q)$\\
    Store TD-Errors in the replay buffer: $p_i = (\abs{\delta_i}+\epsilon)^{\alpha}$\\
    Apply importance sampling $\delta_i = \delta_i w_i$\\
    Update critic by minimizing the huber loss: $L = \frac{1}{N} \sum_i \begin{cases} (\delta_i)^2 & \mathrm{if} \delta < 1\\ \abs{\delta_i} & \mathrm{else}\end{cases}$\\
    Sample a random minibatch of $N$ states $(s_i)$ from $\mathcal{R}$ uniformly\\
    Update the actor policy using the sampled policy gradient: $\Delta_{\theta^{\mu}} J \approx \frac{1}{N} \sum_{i} \Delta_a Q(s_i, a;\theta^Q | a=\mu(s_i)) \Delta_{\theta^{\mu}}\mu(s_i;\theta^{\mu})$\\
    Update the target networks:\\
    $\theta^{Q'} \leftarrow \tau \theta^{Q} + (1-\tau)\theta^{Q'}$\\
    $\theta^{\mu'} \leftarrow \tau \theta^{\mu} + (1-\tau)\theta^{\mu'}$\\
    On death condition, reset the environment
  }
}
\caption{Our DDPG algorithm}
\label{alg:ourddpg}
\end{algorithm}

@@

Each of the controllers reaches close to theoretical optimum and is able to hold it for a number of epochs. With learning rate 1e-3, we reach it somewhere around episode 10, with learning rate 1e-4 a bit later around epoch 25. The time we are holding it before divergence also raises with learning rate. The most aggressive learning rate reached 1828.8 average epoch reward between epoch 7 and 11 and a single maximum of 1917.5 points. On the lower learning rate, it held an average of 1896 points from episode 30 to 56 and reached a maximum of 1968.4 points. The maximum we could reach with our hand-designed PID controller was 1981 points, where the 19 points were lost through the swing at the beginning and then rotational speed was held perfectly.

@@

@@

At the end, most controllers converged to leaving pitch at zero and left the turbine to die regularily with maximum generator torque.

\section{Hold power}
The next task

\section{Pendulum}
To exclude the possibility that our implementation is buggy, we decided to run our DDPG algorithm on the Pendulum task. With our hyperparameters, we could not observe convergence and only switching back to the algorithm of \ac{DDPG}, the agent managed to balance out the pendulum after 11 epochs. Especially the network size hyperparameter was important for this.

@@

Commonly, algorithms are evaluated over many different problems. The OpenAI Gym suite is the common suite for evaluating reinforcement learning algorithms, and typical problems in this are balance problems (all cartpole* and pendulum* tasks) and actuation tasks (gripper*, reacher*, cheetah*). Balance tasks could be argued to resemble our windturbine in a way, as we want to balance an output aswell. However a windturbine is vastly more complex than a pendulum. It exhibits resonant frequencies in addition to the optimization problem which the controller needs to keep under control. Our experiments showed that it didn't. We were able to solve easy tasks, in which resonances of high magnitude still lead to acceptable results, however as soon as it got more difficult, high swinging lead to our control task not being accomplished. Also, noisy sensor measurements were a more difficult task to learn. We could not implement the same size of network and thus not achieve the same learning potential as the authors in \ac{DDPG}.

@@

It was limiting to perform all our experiment on the tiny machine. Limiting factor in this is currently the simulation, even with the lowered accuracy, which currently uses 70\% of the computation time, while we only use 30\% to train our networks. As currently, linux hardware acceleration is broken in qblade, a nice idea could be to fix this and to run future experiments on a cluster. This would also allow efficient hyperparameter sweeps, as we saw that hyperparameter choice greatly affects the outcome of the simulation.

This would also allow to evaluate different wind scenarios and simulations with enabled wakes.

\section{Validate \ac{PER}}

We did not validate our combination of \ac{PER} and \ac{DDPG}. For our task, it gave better results, but it would be interesting to know whether our algorithm might also yield better results on general reinforcement learning tasks. This might also yield some insights into what makes our case special in comparison to general reinforcement learning.

@@

\section{Active control elements}
There is ongoing research into adding active control elements to windturbine blades such as flaps. A difficulty incurred in that research is the design of a controller, as the high-dimensional action space makes it difficult for a human to engineer a controller. Using our automatically learning agent could enable a first working version to that research, providing a first direction or even lead to performance increases over not using active control elements in a turbine.

>>>>

Machine learning has developped immensly over the last years. Benefitting from Moore's law, huge amounts of computational power can be used to train more and more complex models. Machine learning is generally split into three big branches. Supervised learning, unsupervised learning and reinforcement learning. In supervised learning, a function approximator is fitted to model properties of a dataset. In unsupervised learning, properties in a dataset are measured, compared or grouped. In reinforcement learning, the dataset is explored interactively. We are trying to learn a controller for a windturbine, so in theory we could use both supervised an reinforcement learning. If we were to use supervised learning, we would need some precomputed data, which could only be generated by already running a windturbine with a controller and measuring the data. As we however want to learn this controller, it is not present at the start of supervised learning, and imitating known controllers could at best get exactly as good as the known controller. So we are using reinforcement learning.

With the combination of both, we aim to achieve insights into the applicability of reinforcement learning to wind turbine control tasks. We are aiming at laying a foundational work, not necessarily yet outperforming state of the art control systems. Our hope is that the rising performance and more sophisticated algorithms are able to solve pending problems in wind energy now or in the near future.

@@

\section{Gym Experiments}
\label{sec:gymexperiments}
At the start of our experimentation, we took off with a well-known problem and a proven implementation. We used the OpenAI-Pendulum task, in which a pendulum in gravity needs to be held upright by applying a force to it. OpenAI Gyms generally serve as benchmarks for reinforcement learning algorithms and most foundational papers and articles evaluate their solutions based on how well they perform in OpenAI Gyms. We managed to find a solution with our DDPG implementation. We hoped that the implementation might yield results on our windturbine environment as well and exchanged the OpenAI gym for qblade.

@@

The authors of \ac{DDPG} trained their networks up to 2.5 million steps, which we can't afford with our computational power, so a simpler network architecture will hopefully also converge faster.

@@

We normalized the data so states, actions and reward usually stayed between 1 and -1. This is a normal technique in machine learning, and we are irritated by the fact that the reinforcement learning community largely made no mention of normalization. At times, rewards were clipped \cite{mnihPlayingAtariDeep2013} and observations were sometimes, especially when working on pixel inputs, cropped to be of square dimensions. Only in 2016, \cite{vanhasseltLearningValuesMany2016} brought up the topic. They propose an adaptive normalization mechanism which integrates into a generic \ac{RL} algorithm and which can normalize all inputs adaptively. We decided to implement a more straight-forward variant. At first, we used observed states and rewards from the random exploration phase to calculate the 5\%, 95\% quartile values. We calculated normalization constants from it to linearly normalize those quartiles to $[-1, 1]$. Plain min/max normalization worked less well. As sometimes during random exploration, no extreme values were observed and later extreme values lay far outside of $[-1, 1]$, we instead normalized on replay data of a previous run. Additionally, we added a constant of 3 after normalization because of the phenomenon described in section \ref{sec:death}. We could also imagine manually setting normalization bounds on low-dimensional problems, as human knowledge about the possible state space is usually present. In case when human knowledge about input magnitudes is not present, we recommend the mechanism from \cite{vanhasseltLearningValuesMany2016}, or if you are willing to accept the extra computational effort of running once without normalization, our version.

@@

\subsection{Regarding death conditions}
\label{sec:death}

We were observing many exploding turbines in our experiments, so we had a look at how \ac{DDPG} handles death conditions. In \ac{DQN}, death conditions are accounted for by changing the Q target in the terminal step to be $y = r$ instead of $y = r + \gamma \max_a' Q(s', a')$. In fact, we regard the lack of this in \ac{DDPG} as a bug. To explain why, let's have a look at our Q-function again:
\begin{equation}
Q(s,a) = r(s,a) + \gamma Q(s', a')
\end{equation}

The right term of this equation $\gamma Q(s', a')$ is likely several magnitudes higher than the left term $r(s,a)$. With our $\gamma$ = 0.99, circa 99 times as high as $\int_1^{\infty} 0.99^x \approx 98.5$. Though we punish death conditions severely, a death condition will not exceed 3 times the rewards of normal operations due to our clipping. So if, after a death condition, a good condition is reached, our algorithm would gladly take that penalty. For this reason we introduced the updated death notion

\begin{equation}
Q(s,a) = r(s,a) + \begin{cases} 0& \text{if death}\\ Q(s', a') & \text{else}\end{cases}
\end{equation}

However, now it could be advantageous to die in case of negative Q values. Let's suppose we have a terrible policy which under no circumstances produces positive rewards. Thus, our Q term would be negative. As discussed before, it is likely of high magnitude, so elimiating the high negative Q for a 0 would be a good strategy. Thus we need to make sure the Q term is never negative. We do this by adding the lower clipping bound to rewards after clipping them. So now, instead of $y = r + \gamma Q'(s', \mu'(s'))$ we change it to be $y = r + \gamma Q'(s', \mu'(s')) * (1-d)$ where $d \in {0, 1}$ symbolize the death conditions. We haven't found any literature which does a similar thing with \ac{DDPG}

@@

\For{episode = 1, M}{
  Initialize a random process $\mathcal{N}$ for action exploration\\
  Receive initial observation state $s_1$\\
  \For{t = 1, T}{
    Select action $a_t = \mathrm{clip}(\mu(s_t;\theta^{\mu}) + \mathcal{N}_t)$ according to the current policy and exploration noise\\
    Execute action $a_t$ in $\varepsilon$ and observe $r_t$, $d_t$ and $s_{t+1}$ and normalize these\\
    Store transition with highest priority $(s_t, a_t, r_t, s_{t+1}, d_t, \max_i(p_i))$ in $\mathcal{R}$\\
    Calculate sampling probabilities $P(i) = \frac{p_i}{\sum_j p_j}$\\
    Sample a random minibatch of $N$ transitions $(s_i, a_i, r_i, s_{i+1}, d_i)$ from $\mathcal{R}$ according to $P$\\
    Calculate importance sampling weights $w_i = (N \times P(i))^{-\beta}$ and normalize $ w_i = \frac{w_i}{\max_j w_j}$\\
    Set $y_i = r_i + \gamma Q'(s_{i+1}, \mu'(s_{i+1};\theta^{\mu'});\theta^{Q'}) (1 - d_i)$\\
    Calculate TD-Error $\delta_i = y_i - Q(s_i, a_i;\theta^Q)$\\
    Store TD-Errors in the replay buffer: $p_i = (\abs{\delta_i}+\epsilon)^{\alpha}$\\
    Apply importance sampling $\delta_i = \delta_i w_i$\\
    Update critic by minimizing the huber loss: $L = \frac{1}{N} \sum_i \begin{cases} (\delta_i)^2 & \text{if } \delta < 1\\ \abs{\delta_i} & \text{else}\end{cases}$\\
    Sample a random minibatch of $N$ states $(s_i)$ from $\mathcal{R}$ uniformly\\
    Update the actor policy using the sampled policy gradient: $\Delta_{\theta^{\mu}} J \approx \frac{1}{N} \sum_{i} \Delta_a Q(s_i, a;\theta^Q | a=\mu(s_i)) \Delta_{\theta^{\mu}}\mu(s_i;\theta^{\mu})$\\
    Update the target networks:\\
    $\theta^{Q'} \leftarrow \tau \theta^{Q} + (1-\tau)\theta^{Q'}$\\
    $\theta^{\mu'} \leftarrow \tau \theta^{\mu} + (1-\tau)\theta^{\mu'}$\\
    On death condition, reset the environment
  }
}
\caption{Our DDPG algorithm}
\label{alg:ourddpg}
\end{algorithm}

@@

Each of the controllers reaches close to theoretical optimum and is able to hold it for a number of epochs. With learning rate 1e-3, we reach it somewhere around episode 10, with learning rate 1e-4 a bit later around epoch 25. We also tried 1e-5, but we could not see any convergence with this, so we ommitted the plot. The time we are holding it before divergence also raises with learning rate. The most aggressive learning rate reached 1828.8 average epoch reward between epoch 7 and 11 and a single maximum of 1917.5 points. On the lower learning rate, it held an average of 1896 points from episode 30 to 56 and reached a maximum of 1968.4 points. The maximum we could reach with our hand-designed PID controller was 1981 points, where the 19 points were lost through the swing at the beginning and then rotational speed was held perfectly.

@@

From this we see that higher learning rates cause a more stable learning, but reach convergence later. The results suffer from strong oscillations both with the low and high learning rate.

@@

At the end, most controllers converged to leaving pitch at zero and left the turbine to die regularily with maximum generator torque. Though this is not an optimum in the global reward function, apparently this became an optimum in the Q estimate. Looking back at our explanation from section \ref{sec:death} where we gave some options on why an agent could favor a death over returning to a sensible region, we can exclude the case that good Q values after a death (because after a reset the agent returns to sensible area) propagate. So we looked at the case of negative Q estimates. When plotting Q estimates, we found that indeed Q is sometimes predicting negative values, especially in early stages of training.

A possible explanation for favouring deaths could be that our death penalties are spikes in the reward functions. As explained in section \ref{sec:rewardfunctions}, we add a constant penalty upon death but not in the steps before. As our Q function approximator approximates smooth functions, it will likely have a high TD-Error on these samples. This in turn will cause \ac{PER} to replay these samples relatively often in comparison to other samples. As we didn't use a \ac{PER} beta of 1 to fully correct against importance sampling bias but 0.5 as proposed in the paper, we could imagine that still some oversampling happened along these points and they actually ended up being largely overestimated.

Also, we could imagine that the Q function gets deformed before deaths and maybe actually exposes a local maximum right before it. The spikey nature could lead to weird situations in a high-dimensional regression, which could be similar to overfitting. As the regression wants to fit to the outlier, it might create a strong curvature before this. To illustrate this, let's plot a typical overfit scenario in figure \ref{fig:overfit}. The dot on the bottom right is our spike in death. Deriving along the linear regression (blue line) would lead us away from the death. The orange line however exposes a local maximum shortly before the death. If our policy for some reason got stuck in this local maximum, it would be trained to stay close to a death condition, which would, because of system instability, lead to many death conditions happening. Oversampling effects might then even increase this phenomenon.

We are unsure why the controllers showed such a bad performance, and couldn't get to fix this problem anymore.

\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{images/overfit.png}
  \caption{Overfitting a 5 point regression}
  \label{fig:overfit}
\end{figure}

\section{Pendulum}
To exclude the possibility that our implementation is buggy, we decided to run our DDPG algorithm on the Pendulum task. With our hyperparameters, we could not observe convergence and only switching back to the algorithm of \ac{DDPG} which we already tried in \ref{sec:gymexperiments}, the agent managed to balance out the pendulum. Thus we can not exclude misimplementation as a bug might only occur with our set of hyperparameters, however we suspect that windturbines are just not very similar to pendulums.

@@

To summarize it, the level of control we have achieved does not outperform any previous benchmark controller. In fact, it would destroy most windturbines relatively quickly and does not yield sensible energy. Still we showed that it is possible, that existing algorithms however exhibit some significant problems and need reworking. Also we showed that we are not able to perform this reworking in the given amount of time.

Commonly, algorithms are evaluated over many different problems. The OpenAI Gym suite is the common suite for evaluating reinforcement learning algorithms, and typical problems in this are balance problems (all cartpole* and pendulum* tasks) and actuation tasks (gripper*, reacher*, cheetah*). Balance tasks could be argued to resemble our windturbine in a way, as we want to balance an output aswell. However a windturbine is apparently different enough so we can not optimize it with the same algorithm which optimized the other games. It exhibits resonant frequencies in addition to the optimization problem which the controller needs to keep under control. Our experiments showed that it didn't. We were able to solve easy tasks, in which resonances of high magnitude still lead to acceptable results, however as soon as it got more difficult, high swinging lead to our control task not being accomplished. Also, noisy sensor measurements were a more difficult task to learn. We could not implement the same size of network and thus not achieve the same learning potential as the authors in \ac{DDPG}.

@@

It was limiting to perform all our experiment on the tiny machine. Limiting factor in this is currently the simulation, even with the lowered accuracy, which currently uses 70\% of the computation time, while we only use 30\% to train our networks. As currently, linux hardware acceleration is broken in qblade, a nice idea could be to fix this and to run future experiments on a cluster. This would also allow efficient hyperparameter sweeps, as we saw that hyperparameter choice greatly affects the outcome of the simulation. Also, it would be interesting to run simulations for a longer time. The authors of \ac{TD3} ran their simulations for up to 1 million steps, \ac{DDPG} 2.5 million steps and \ac{DQN} 50 million. On easier tasks, TD3 could reach first sensible results after 0.2 million steps. With our computational power we couldn't run anything above 0.3 million steps. But we also chose a less complex architecture which should be faster to train.

With high computational power, we could also evaluate many different wind scenarios and turbine models, in contrast to our evaluation which was done on a single wind speed and turbine type.

\section{Validate against OpenAI-Gyms}

We did not validate our combination of \ac{PER} and \ac{DDPG}. Neither did we validate our addition of death conditions. For our task, it gave better results, but it would be interesting to know whether our algorithm might also yield better results on general reinforcement learning tasks. Running through the suite of OpenAI Gyms would be an interesting work and could show whether our critique on \ac{DDPG} was justified. Also, we might be able to get \ac{TD3} to work there. This comparison might also yield some insights into what makes our case special in comparison to general reinforcement learning.

@@

\section{Active control elements}
There is ongoing research into adding active control elements to windturbine blades such as flaps. A difficulty incurred in that research is the design of a controller, as the high-dimensional action space makes it difficult for a human to engineer a controller. As soon as we could provide a working version of an automatically learning, stable agent, we could assist the efforts of the people working on flap design to automatically learn controllers for them.
